---
title: "Advanced Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>'
)
```

This vignette covers advanced `irelink` features for users who have
completed the Getting Started or Deduplication vignettes.
All examples use `fake_1000` and a shared in-memory DuckDB connection.

## Setup

Train a complete model on `fake_1000` that will be reused across sections.

```{r setup}
library(irelink)
library(ggplot2)

df <- fake_1000
con <- DBI::dbConnect(duckdb::duckdb())

spec <- il_spec() |>
  il_compare(first_name, cl_name()) |>
  il_compare(surname, cl_name()) |>
  il_compare(dob, cl_dob()) |>
  il_compare(city, cl_exact(term_frequency = TRUE)) |>
  il_compare(email, cl_email()) |>
  il_block_on(first_name) |>
  il_block_on(surname) |>
  il_block_on(city)

model <- il_model(df, spec = spec, con = con)
model <- il_estimate_prior(
  model,
  block_on(first_name, surname),
  block_on(email),
  recall = 0.6
)
model <- il_estimate_u(model, max_pairs = 1e5)
model <- il_estimate_em(model, block_on(first_name))
model <- il_estimate_em(model, block_on(dob))

pairs <- predict(model, threshold = 0.5)
clusters <- il_cluster(pairs, threshold = 0.85)
```

## Training diagnostics

`il_training_history()` returns the m and u parameter estimates at each
EM iteration across all training sessions.
Plot it to check whether parameters have converged:

```{r history, fig.width = 7, fig.height = 5}
hist <- il_training_history(model)
autoplot(hist)
```

A well-converged model shows stable values in the final iterations.
If estimates are still drifting, add more EM passes with different
blocking rules or expand the set of candidate pairs.

## Pair inspection

`il_compare_records()` scores a single pair of records against the
spec without requiring a full prediction pass.
Use it to diagnose why a known match scores too low or a non-match
scores too high.

```{r compare-records}
rec_a <- fake_1000[1, ]
rec_b <- fake_1000[5, ]

il_compare_records(rec_a, rec_b, spec = model$spec, con = con)
```

The gamma columns show the comparison level reached on each field.
Cross-reference these with `il_weights(model)` to read off the
match-weight contribution of each level.

For a visual breakdown of how per-field gamma values sum to an overall
match decision, draw a waterfall chart for any pair in the scored set:

```{r waterfall, fig.width = 6, fig.height = 3}
autoplot(pairs, which = 1)
```

## Lazy prediction for large data

`predict(collect = FALSE)` keeps scored pairs in the database rather
than collecting them into R.
`il_cluster()` detects the lazy reference and runs connected-components
analysis entirely in SQL, avoiding a costly round-trip for datasets
where materialising millions of rows would exhaust memory.

```{r lazy}
pairs_lazy <- predict(model, threshold = 0.5, collect = FALSE)
pairs_lazy
```

Pass the lazy reference directly to `il_cluster()`:

```{r lazy-cluster}
clusters_lazy <- il_cluster(pairs_lazy, threshold = 0.85)
nrow(clusters_lazy)
```

`autoplot()` and `il_waterfall()` collect automatically when needed,
so downstream analysis code is unchanged.
Use the lazy path any time candidate-pair counts exceed available memory.

## Cluster diagnostics

`il_graph_metrics()` computes node-, edge-, and cluster-level summaries
from the linkage graph.
Use it to spot over-generous thresholds (clusters that are too large)
or sparse connectivity (clusters with unexpectedly low density).

```{r graph-metrics}
metrics <- il_graph_metrics(pairs, clusters)
```

The cluster table reports size and internal edge density
(edges / maximum possible edges for a cluster of that size):

```{r graph-clusters}
metrics$clusters
```

A high maximum cluster size combined with low density often indicates
that a transitive link is pulling unrelated entities together.
Consider raising the threshold in `predict()` or `il_cluster()` and
re-checking the metrics.

The node table shows how many links each record participates in (degree):

```{r graph-nodes}
head(metrics$nodes)
```

Records with unusually high degree relative to their cluster size may
be acting as hubs that inflate the cluster beyond its true membership.

## Phonetic blocking

Standard equality blocking misses pairs where names are spelled
differently but sound alike — for example, "Smith" / "Smyth" or
"Jon" / "John".
Pass `.transform = il_soundex` to `il_block_on()` or `block_on()` to
group names by phonetic code instead of exact spelling.

```{r phonetic-spec}
spec_phon <- il_spec() |>
  il_compare(first_name, cl_name()) |>
  il_compare(surname, cl_name()) |>
  il_compare(dob, cl_dob()) |>
  il_block_on(first_name, .transform = il_soundex) |>
  il_block_on(surname, .transform = il_soundex)
```

Use the same `.transform` argument when specifying the blocking rule
for an EM training pass:

```{r phonetic-train}
model_phon <- il_model(df, spec = spec_phon, con = con)
model_phon <- il_estimate_u(model_phon, max_pairs = 1e5)
model_phon <- il_estimate_em(
  model_phon,
  block_on(first_name, .transform = il_soundex)
)
```

Phonetic blocking increases recall at the cost of more candidate pairs.
Use `il_count_pairs()` to check the volume trade-off before committing
to a spec.

## Column transforms

The `transform` argument in `il_compare()` applies a function to both
values before scoring.
Use it to normalise case or strip whitespace before a similarity
comparison:

```{r transform-spec}
spec_tr <- il_spec() |>
  il_compare(first_name, cl_jaro_winkler(0.9, 0.7), transform = tolower) |>
  il_compare(surname, cl_jaro_winkler(0.9, 0.7), transform = tolower) |>
  il_compare(dob, cl_exact()) |>
  il_block_on(first_name) |>
  il_block_on(surname)

model_tr <- il_model(df, spec = spec_tr, con = con)
model_tr <- il_estimate_u(model_tr, max_pairs = 1e5)
model_tr <- il_estimate_em(model_tr, block_on(surname))
```

`tolower`, `toupper`, and `trimws` are translated to SQL on DuckDB and
PostgreSQL, so the transform runs in-database at full speed.
Custom R functions work on the R-side path only.
When saving a model with `il_save()`, `.rds` keeps the R object as-is.
`.json` writes Splink settings SQL, so loaded comparisons come back as
SQL-backed levels. Anonymous functions still produce a warning on save.

## Incremental matching

`il_find_matches()` scores new records against the data already loaded
into a trained model without retraining.
It applies the same blocking rules and comparison spec:

```{r find-matches}
new_df <- data.frame(
  first_name = c('Jhon', 'Alice'),
  surname    = c('Smith', 'Jones'),
  dob        = c('1990-01-15', '1985-06-20'),
  city       = c('London', 'Manchester'),
  email      = c(NA, 'ajones@example.com')
)

matches <- il_find_matches(model, new_df, threshold = 0.5)
matches
```

Each row is a (new record, existing record) pair.
`unique_id_l` identifies the new record (auto-assigned starting from 1)
and `unique_id_r` identifies the matched record in the original dataset.

This pairs naturally with `il_load()` and `il_attach()`: load a saved
model, attach it to the current database, and call `il_find_matches()`
for each incoming batch of new records.

## Cleanup

```{r cleanup}
il_cleanup(model)
il_cleanup(model_phon)
il_cleanup(model_tr)
DBI::dbDisconnect(con, shutdown = TRUE)
```
