---
title: "Deduplication with Evaluation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deduplication with Evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette walks through a complete deduplication workflow on the
`fake_1000` dataset, including model training, prediction, clustering,
and evaluation against ground-truth labels.
The dataset is the primary demo data from the Python
[splink](https://github.com/moj-analytical-services/splink) library.
It contains 1,000 records representing 181 unique people, each
with varying numbers of duplicate entries corrupted with typos, missing
values, and other realistic data-quality issues.

## Setup

```{r setup}
library(irelink)
library(ggplot2)
```

## Explore the data

```{r load-data}
df <- fake_1000
head(df)
```

The `cluster` column is the ground truth: records sharing the same
cluster value refer to the same person.
There are `r length(unique(df$cluster))` unique entities across
`r nrow(df)` records.
Note that missing values appear as `NA`, reflecting real-world
data-quality challenges.

Before building a model, profile the data to understand its completeness
and value distributions:

```{r completeness}
con <- DBI::dbConnect(duckdb::duckdb())
comp <- il_completeness(df, con = con)
comp
```

```{r completeness-plot, fig.width = 6, fig.height = 3}
autoplot(comp)
```

Check column value distributions to inform blocking and comparison
choices:

```{r profile}
il_profile(df[, c("first_name", "surname", "city")], con = con, top_n = 5)
```

## Define the specification

Choose comparisons and blocking rules.
Names use Jaro-Winkler similarity, dates of birth use the `cl_dob()`
helper, and city uses exact matching with term-frequency adjustments:

```{r spec}
spec <- il_spec() |>
  il_compare(first_name, cl_name()) |>
  il_compare(surname, cl_name()) |>
  il_compare(dob, cl_dob()) |>
  il_compare(city, cl_exact(term_frequency = TRUE)) |>
  il_compare(email, cl_email()) |>
  il_block_on(first_name) |>
  il_block_on(surname) |>
  il_block_on(city)

spec
```

Estimate how many pairs each blocking rule generates:

```{r count-pairs}
il_count_pairs(
  df,
  block_on(first_name),
  block_on(surname),
  block_on(city),
  con = con
)
```

## Train the model

```{r model}
model <- il_model(df, spec = spec, con = con)
```

Estimate the prior match probability using deterministic rules:

```{r prior}
model <- il_estimate_prior(
  model,
  block_on(first_name, surname),
  block_on(email),
  recall = 0.6
)
```

Estimate u-probabilities from random pairs and m-probabilities via EM:

```{r train}
model <- il_estimate_u(model, max_pairs = 1e5)
model <- il_estimate_em(model, block_on(first_name))
model <- il_estimate_em(model, block_on(dob))
```

## Inspect the trained model

```{r summary}
summary(model)
```

The match weights chart shows the discriminative power of each
comparison:

```{r weights-plot, fig.width = 6, fig.height = 3.5}
autoplot(model)
```

The parameter chart shows the raw m and u probabilities:

```{r params-plot, fig.width = 7, fig.height = 4}
autoplot(model, type = "parameters")
```

## Predict and cluster

Score all candidate pairs and apply a probability threshold:

```{r predict}
predictions <- predict(model, threshold = 0.5)
nrow(predictions)
```

View the match-weight distribution:

```{r histogram, fig.width = 6, fig.height = 3}
autoplot(predictions)
```

Inspect how individual pairs are scored with a waterfall chart:

```{r waterfall, fig.width = 6, fig.height = 3}
autoplot(predictions, which = 1)
```

Resolve pairwise links into entity clusters:

```{r cluster}
clusters <- il_cluster(predictions, threshold = 0.85)
head(clusters)
```

## Evaluate against ground truth

The `cluster` column in the original data provides ground-truth entity
labels.
Convert these to pairwise labels for evaluation:

```{r labels}
# Use the bundled clerical labels from splink
labels_raw <- fake_1000_labels

# Rename to match irelink's evaluation convention
labels <- data.frame(
  unique_id_l = labels_raw$unique_id_l,
  unique_id_r = labels_raw$unique_id_r,
  is_match = as.integer(labels_raw$clerical_match_score)
)

nrow(labels)
sum(labels$is_match)
```

### Accuracy metrics

```{r accuracy}
acc <- il_accuracy(model, labels = labels)
acc
```

```{r accuracy-plot, fig.width = 6, fig.height = 3.5}
autoplot(acc)
```

### ROC curve

```{r roc, fig.width = 5, fig.height = 4}
roc <- il_roc(model, labels = labels)
autoplot(roc)
```

### Precision–recall curve

```{r pr, fig.width = 5, fig.height = 4}
pr <- il_precision_recall(model, labels = labels)
autoplot(pr)
```

### Error inspection

Examine false positives and false negatives at a specific threshold:

```{r errors}
errors <- il_errors(model, labels = labels, threshold = 0.85)
head(errors)
```

### Unlinkables

How many records cannot be linked at each threshold?

```{r unlinkables, fig.width = 6, fig.height = 3}
unlink <- il_unlinkables(model)
autoplot(unlink)
```

## Cleanup

```{r cleanup}
il_cleanup(model)
DBI::dbDisconnect(con, shutdown = TRUE)
```
