---
title: "Record Linkage Across Datasets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Record Linkage Across Datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>'
)
```

This vignette demonstrates linking records across two separate datasets
using the FEBRL (Freely Extensible Biomedical Record Linkage)
benchmark data.
Dataset 4a contains 5,000 original records and dataset 4b contains
5,000 duplicates — one for each original — corrupted with
typographical errors, missing values, and transpositions.
Ground truth is encoded in the `rec_id` column: records sharing the
same base number (e.g., `rec-1070-org` and `rec-1070-dup-0`) refer to
the same entity.

## Setup

```{r setup}
library(irelink)
library(ggplot2)
```

## Load the data

```{r head}
head(febrl4a)
head(febrl4b)
```

## Explore data quality

Check completeness across both tables.
Table B has more missing values due to the corruption process:

```{r completeness}
con <- DBI::dbConnect(duckdb::duckdb())
comp <- il_completeness(febrl4a, febrl4b, con = con)
comp
```

```{r completeness-plot, fig.width = 6, fig.height = 3.5}
autoplot(comp)
```

## Define the specification

For linking (as opposed to deduplication), set `link_type = "link"` when
creating the model.
Comparisons use name similarity, date-of-birth matching, and postcode
exact matching:

```{r spec}
spec <- il_spec() |>
  il_compare(given_name, cl_name()) |>
  il_compare(surname, cl_name()) |>
  il_compare(date_of_birth, cl_exact()) |>
  il_compare(postcode, cl_exact()) |>
  il_block_on(surname) |>
  il_block_on(given_name)

spec
```

## Train the model

Create a link-type model with both tables:

```{r model}
model <- il_model(
  febrl4a, febrl4b,
  spec = spec,
  con = con,
  link_type = 'link'
)

model
```

Estimate prior match probability, u-probabilities, and run EM:

```{r train}
model <- il_estimate_prior(
  model,
  block_on(given_name, surname),
  block_on(surname, suburb),
  recall = 0.6
)

model <- il_estimate_u(model, max_pairs = 1e5)
model <- il_estimate_em(model, block_on(surname))
model <- il_estimate_em(model, block_on(suburb))
```

## Inspect the model

```{r weights, fig.width = 6, fig.height = 3.5}
autoplot(model)
```

```{r params, fig.width = 7, fig.height = 4}
autoplot(model, type = 'parameters')
```

```{r weights-table}
il_weights(model)
```

## Predict and cluster

```{r predict}
predictions <- predict(model, threshold = 0.5)
nrow(predictions)
```

```{r hist, fig.width = 6, fig.height = 3}
autoplot(predictions)
```

Cluster the pairs to resolve entities:

```{r cluster}
clusters <- il_cluster(predictions, threshold = 0.85)
head(clusters)
```

## Evaluate against ground truth

The `rec_id` column encodes ground truth.
Extract entity IDs and build pairwise labels:

```{r labels}
# Extract entity number from rec_id (e.g., "rec-1070-org" -> "1070")
entity_a <- sub('^rec-(\\d+)-org$', '\\1', febrl4a$rec_id)
entity_b <- sub('^rec-(\\d+)-dup-\\d+$', '\\1', febrl4b$rec_id)

# Build id-entity lookup (unique_id auto-generated by il_model)
ids_a <- data.frame(unique_id = seq_len(nrow(febrl4a)), entity = entity_a)
ids_b <- data.frame(unique_id = seq_len(nrow(febrl4b)), entity = entity_b)

# True matches: same entity across tables
positives <- merge(ids_a, ids_b, by = 'entity')
names(positives) <- c('entity', 'unique_id_l', 'unique_id_r')
positives$is_match <- 1L
positives <- positives[, c('unique_id_l', 'unique_id_r', 'is_match')]

# Sample non-matching pairs
set.seed(42)
n_neg <- min(nrow(positives), 2000L)
neg_l <- sample(ids_a$unique_id, n_neg, replace = TRUE)
neg_r <- sample(ids_b$unique_id, n_neg, replace = TRUE)
ent_l <- ids_a$entity[match(neg_l, ids_a$unique_id)]
ent_r <- ids_b$entity[match(neg_r, ids_b$unique_id)]
negatives <- data.frame(
  unique_id_l = neg_l,
  unique_id_r = neg_r,
  is_match = ifelse(ent_l == ent_r, 1L, 0L)
)

labels <- rbind(positives, negatives)
nrow(labels)
sum(labels$is_match)
```

### Accuracy metrics

```{r accuracy, fig.width = 6, fig.height = 3.5}
acc <- il_accuracy(model, labels = labels)
autoplot(acc)
```

### ROC and Precision–Recall

```{r roc, fig.width = 5, fig.height = 4}
roc <- il_roc(model, labels = labels)
autoplot(roc)
```

```{r pr, fig.width = 5, fig.height = 4}
pr <- il_precision_recall(model, labels = labels)
autoplot(pr)
```

## Cleanup

```{r cleanup}
il_cleanup(model)
DBI::dbDisconnect(con, shutdown = TRUE)
```
