---
title: "Linking Banking Transactions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linking Banking Transactions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
can_run <- requireNamespace('nanoparquet', quietly = TRUE)

if (can_run) {
  base_url <- paste0(
    'https://raw.githubusercontent.com/',
    'moj-analytical-services/splink_datasets/',
    'master/data/'
  )
  tmp_ori <- paste0(tempdir(), '/transactions_origin.parquet')
  tmp_dest <- paste0(tempdir(), '/transactions_destination.parquet')
  df_origin <- try(
    {
      utils::download.file(
        paste0(base_url, 'transactions_origin.parquet'),
        tmp_ori,
        mode = 'wb',
        quiet = TRUE
      )
      nanoparquet::read_parquet(tmp_ori)
    },
    silent = TRUE
  )
  df_destination <- try(
    {
      utils::download.file(
        paste0(base_url, 'transactions_destination.parquet'),
        tmp_dest,
        mode = 'wb',
        quiet = TRUE
      )
      nanoparquet::read_parquet(tmp_dest)
    },
    silent = TRUE
  )
  can_run <- !inherits(df_origin, 'try-error') &&
    !inherits(df_destination, 'try-error')
}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>',
  eval = can_run
)
```

This vignette replicates the [Splink "Linking banking transactions" demo](https://moj-analytical-services.github.io/splink/demos/examples/duckdb/transactions.html) using `irelink`.
It demonstrates two-table linking (`link_type = "link"`), matching each outgoing payment in the origin table to its corresponding incoming payment in the destination table.

The data is synthetic and designed to be challenging: amounts differ due to fees and exchange-rate effects, dates shift by a few days, and memos are sometimes truncated.
Since every origin payment has exactly one destination counterpart, the prior match probability is `1 / n_origin`.

This vignette requires the [nanoparquet](https://cran.r-project.org/package=nanoparquet) package to read the remote Parquet files and will only compile when the package and data URLs are both available.

## Load the data

```{r load-data}
library(irelink)
library(ggplot2)

df_origin
df_destination
```

## Profile the data

```{r setup-con}
con <- DBI::dbConnect(duckdb::duckdb())
```

```{r profile}
il_profile(df_origin, memo, transaction_date, amount, con = con, top_n = 8)
```

## Choose blocking rules

Because corresponding records differ systematically — amounts change due
to fees, dates shift, memos are truncated — blocking rules must be
generous enough to capture true matches while still reducing the pair
space dramatically.
The rules below use SQL expressions via `.where`:

```{r count-pairs}
counts <- il_count_pairs(
  df_origin,
  df_destination,
  # Same year-month, similar memo prefix, amount ratio within 30%
  block_on(
    .where = paste(
      "strftime(l.transaction_date, '%Y%m') = strftime(r.transaction_date, '%Y%m')",
      'AND substr(l.memo, 1, 3) = substr(r.memo, 1, 3)',
      'AND l.amount / r.amount > 0.7 AND l.amount / r.amount < 1.3'
    )
  ),
  # Same but offset by 15 days to catch month boundaries
  block_on(
    .where = paste(
      "strftime(l.transaction_date + 15, '%Y%m') = strftime(r.transaction_date, '%Y%m')",
      'AND substr(l.memo, 1, 3) = substr(r.memo, 1, 3)',
      'AND l.amount / r.amount > 0.7 AND l.amount / r.amount < 1.3'
    )
  ),
  # Memo prefix (first 9 characters)
  block_on(.where = 'substr(l.memo, 1, 9) = substr(r.memo, 1, 9)'),
  # Rounded amount + same week
  block_on(
    .where = paste(
      'round(l.amount / 2, 0) * 2 = round(r.amount / 2, 0) * 2',
      'AND yearweek(r.transaction_date) = yearweek(l.transaction_date)'
    )
  ),
  # Amount offset + week offset
  block_on(
    .where = paste(
      'round(l.amount / 2, 0) * 2 = round((r.amount + 1) / 2, 0) * 2',
      'AND yearweek(r.transaction_date) = yearweek(l.transaction_date + 4)'
    )
  ),
  # Ground-truth "cheat" rule for completeness
  block_on(unique_id),
  con = con,
  link_type = 'link'
)
counts
```

```{r count-pairs-plot, fig.width = 7, fig.height = 4}
autoplot(counts)
```

## Define the specification

The `transaction_date` comparison is one-sided: payments can only *arrive* after they are *sent*, so the comparison checks `destination_date - origin_date` is between 0 and *N* days:

```{r spec}
spec <- il_spec() |>
  il_compare(amount, cl_pct_diff(0.01, 0.03, 0.10, 0.30)) |>
  il_compare(memo, cl_levenshtein(2, 6, 10)) |>
  il_compare(
    transaction_date,
    cl_levels(
      cl_null(),
      cl_custom('(r.{col} - l.{col}) BETWEEN 0 AND 1'),
      cl_custom('(r.{col} - l.{col}) BETWEEN 0 AND 4'),
      cl_custom('(r.{col} - l.{col}) BETWEEN 0 AND 10'),
      cl_custom('(r.{col} - l.{col}) BETWEEN 0 AND 30'),
      cl_else()
    )
  ) |>
  il_block_on(
    .where = paste(
      "strftime(l.transaction_date, '%Y%m') = strftime(r.transaction_date, '%Y%m')",
      'AND substr(l.memo, 1, 3) = substr(r.memo, 1, 3)',
      'AND l.amount / r.amount > 0.7 AND l.amount / r.amount < 1.3'
    )
  ) |>
  il_block_on(
    .where = paste(
      "strftime(l.transaction_date + 15, '%Y%m') = strftime(r.transaction_date, '%Y%m')",
      'AND substr(l.memo, 1, 3) = substr(r.memo, 1, 3)',
      'AND l.amount / r.amount > 0.7 AND l.amount / r.amount < 1.3'
    )
  ) |>
  il_block_on(.where = 'substr(l.memo, 1, 9) = substr(r.memo, 1, 9)') |>
  il_block_on(
    .where = paste(
      'round(l.amount / 2, 0) * 2 = round(r.amount / 2, 0) * 2',
      'AND yearweek(r.transaction_date) = yearweek(l.transaction_date)'
    )
  ) |>
  il_block_on(
    .where = paste(
      'round(l.amount / 2, 0) * 2 = round((r.amount + 1) / 2, 0) * 2',
      'AND yearweek(r.transaction_date) = yearweek(l.transaction_date + 4)'
    )
  ) |>
  il_block_on(unique_id)

spec
```

## Train the model

```{r model}
model <- il_model(
  df_origin,
  df_destination,
  spec = spec,
  con = con,
  link_type = 'link'
)
model$params$prior <- 1 / nrow(df_origin)
model <- il_estimate_u(model, max_pairs = 1e6) |>
  il_estimate_em(block_on(memo)) |>
  il_estimate_em(block_on(amount))
```

## Inspect the trained model

```{r summary}
summary(model)
```

```{r weights-plot, fig.width = 7, fig.height = 4}
autoplot(model)
```

```{r params-plot, fig.width = 7, fig.height = 5}
autoplot(model, type = 'parameters')
```

## Predict

```{r predict}
predictions <- predict(model, threshold = 0.001)
predictions
```

```{r histogram, fig.width = 7, fig.height = 3.5}
autoplot(predictions)
```

```{r waterfall, fig.width = 7, fig.height = 4}
autoplot(predictions, which = 1)
```

## Evaluate against ground truth

```{r accuracy}
acc <- il_accuracy(model, labels_col = 'ground_truth')
acc
```

```{r accuracy-plot, fig.width = 7, fig.height = 4}
autoplot(acc)
```

### Error inspection

```{r errors-fp}
errors <- il_errors(model, labels_col = 'ground_truth', threshold = 0.5)
errors[errors$error_type == 'false_positive', ]
```

```{r errors-fn}
errors[errors$error_type == 'false_negative', ]
```

## Cleanup

```{r cleanup}
il_cleanup(model)
DBI::dbDisconnect(con, shutdown = TRUE)
```

`il_cleanup(model)` is model-scoped. If an interactive run failed before you
kept the model object, call `il_cleanup_all(con)` to remove all `irelink`
tables from the connection before disconnecting.
