---
title: "Accuracy benchmark: FEBRL 4 record linkage"
format:
  html:
    embed-resources: true
---

This document reproduces the [splink FEBRL 4 demo](https://moj-analytical-services.github.io/splink/demos/examples/duckdb/febrl4.html) using irelink and evaluates accuracy against the ground truth encoded in the data.
FEBRL 4 is a standard record-linkage benchmark: `febrl4a` contains 5,000 original records and `febrl4b` contains 5,000 corrupted duplicates, one per original, with typographical errors, missing values, and transpositions.
Because the match structure is one-to-one, the full set of true positive pairs is known.

```{r setup}
#| message: false
devtools::load_all()
library(dplyr)
library(ggplot2)
library(stringr)

con <- DBI::dbConnect(duckdb::duckdb())
```

## Prepare the data

The FEBRL generator stores `date_of_birth`, `soc_sec_id`, `postcode`, and `street_number` as integers.
We convert them to character so that string-distance comparisons (Damerau-Levenshtein) work correctly.

```{r prepare}
prep <- function(df) {
  mutate(
    df,
    across(c(date_of_birth, soc_sec_id, postcode, street_number), as.character)
  )
}

df_a <- prep(febrl4a)
df_b <- prep(febrl4b)
```

## Explore data quality

Check how complete each column is across both tables.
Table B has more missing values because the corruption process drops fields.

```{r completeness}
#| fig-width: 7
#| fig-height: 3.5
il_completeness(df_a, df_b, con = con) |> autoplot()
```

Profile the highest-frequency values for the name columns.
Highly skewed distributions inflate u-probabilities, so spotting them early is useful.

```{r profile}
#| fig-width: 7
#| fig-height: 3.5
il_profile(df_a, given_name, surname, con = con, top_n = 8) |> autoplot()
```

## Blocking analysis

We use seven blocking rules, matching the splink tutorial.
Together they generate around 55,000 candidate pairs from 25 million possible cross-table combinations, a reduction of roughly 450x.

```{r count-pairs}
#| fig-width: 7
#| fig-height: 3.5
il_count_pairs(
  df_a,
  df_b,
  block_on(given_name, surname),
  block_on(.where = 'l.given_name = r.surname AND l.surname = r.given_name'),
  block_on(date_of_birth),
  block_on(soc_sec_id),
  block_on(state, address_1),
  block_on(street_number, address_1),
  block_on(postcode),
  con = con,
  link_type = 'link'
) |>
  autoplot()
```

## Define the specification

The comparisons follow the splink detailed model: fuzzy name matching, date-of-birth and social-security-ID edit distance, exact street number, and postcode edit distance.
Term-frequency adjustment is applied to street number to avoid inflating match weights for common values like "1".

```{r spec}
spec <- il_spec() |>
  il_compare(col = given_name, method = cl_name()) |>
  il_compare(col = surname, method = cl_name()) |>
  il_compare(col = date_of_birth, method = cl_damerau_levenshtein(1, 2)) |>
  il_compare(col = soc_sec_id, method = cl_damerau_levenshtein(1, 2)) |>
  il_compare(col = street_number, method = cl_exact(term_frequency = TRUE)) |>
  il_compare(col = postcode, method = cl_damerau_levenshtein(1, 2)) |>
  il_block_on(given_name, surname) |>
  il_block_on(
    .where = 'l.given_name = r.surname AND l.surname = r.given_name'
  ) |>
  il_block_on(date_of_birth) |>
  il_block_on(soc_sec_id) |>
  il_block_on(state, address_1) |>
  il_block_on(street_number, address_1) |>
  il_block_on(postcode)

spec
```

## Train the model

We build a link-type model across two source tables, then estimate parameters in three stages:


1. **Prior:** the baseline match probability, estimated from deterministic rules assumed to capture about 80% of true matches.
2. **U-probabilities:** the chance of agreement by coincidence, estimated from random record pairs.
3. **M-probabilities:** the chance of agreement among true matches, estimated by two EM sessions that each hold a different blocking rule fixed.

```{r train}
model <- il_model(
  .data = df_a,
  df_b,
  spec = spec,
  con = con,
  link_type = 'link'
) |>
  il_estimate_prior(
    block_on(soc_sec_id),
    block_on(given_name, surname, date_of_birth),
    recall = 0.8
  ) |>
  il_estimate_u(max_pairs = 1e6) |>
  il_estimate_em(blocking = block_on(date_of_birth)) |>
  il_estimate_em(blocking = block_on(postcode))

model
```

## Inspect parameters

The match-weight chart summarises the log-Bayes-factor each comparison contributes.
`soc_sec_id` dominates because even a single-edit match is strong evidence for a field that is nearly unique per person.

```{r weights}
#| fig-width: 6
#| fig-height: 3.5
autoplot(model)
```

The m/u chart separates the two components of each weight.
High m (agreement rate among true matches) combined with low u (agreement rate among random pairs) produces a large match weight.

```{r params}
#| fig-width: 7
#| fig-height: 4
autoplot(model, type = 'parameters')
```

The underlying numbers are available via `il_weights()`, which reports the log-2 Bayes factor and the m and u probabilities for each comparison level.

```{r weights-table}
il_weights(model)
```

The training history shows how m and u estimates evolved across EM iterations.
Flat lines indicate the parameters converged quickly; a jagged path suggests the blocking rule for that session provided little signal.

```{r history}
#| fig-width: 7
#| fig-height: 4
il_training_history(model) |>
  autoplot()
```

## Predict

We generate all candidate pairs above a low threshold of 0.1 to retain near-matches for inspection, then examine the match-weight distribution.
A bimodal shape, with a cluster near zero for non-matches and a cluster at high weights for matches, indicates good model separation.

```{r predict}
#| fig-width: 6
#| fig-height: 3
predictions <- predict(model, threshold = 0.1)
nrow(predictions)
autoplot(predictions)
```

A waterfall chart makes that distribution more interpretable by showing which fields pushed an individual pair up or down.
Here we inspect the predicted pair closest to the `0.9` decision threshold used later in the error analysis.

```{r waterfall-pair}
inspect_idx <- predictions |>
  mutate(
    row_id = row_number(),
    distance_to_review_threshold = abs(match_probability - 0.9)
  ) |>
  slice_min(distance_to_review_threshold, n = 1, with_ties = FALSE) |>
  pull(row_id)

predictions |>
  slice(inspect_idx) |>
  select(unique_id_l, unique_id_r, match_weight, total_match_weight, match_probability)
```

```{r waterfall-plot}
#| fig-width: 7
#| fig-height: 4
autoplot(predictions, which = inspect_idx)
```

## Evaluate

Because every record in `febrl4b` is a corrupted duplicate of exactly one record in `febrl4a`, the full set of 5,000 true positive pairs is known.
We recover them by matching on the entity number embedded in `rec_id` (for example, `rec-1070-org` and `rec-1070-dup-0` share entity `1070`).
The model assigns `unique_id` in row order, so the mapping from data rows to model IDs is direct.
For negatives, we avoid random cross-table pairs because they are usually trivially non-matching and can make ROC and precision-recall curves look artificially perfect.
Instead we sample hard negatives from the blocked candidate-pair space: pairs the model genuinely considered, but which belong to different entities.

```{r labels}
ids_a <- df_a |>
  mutate(
    unique_id = row_number(),
    entity = str_extract(rec_id, '(?<=rec-)\\d+')
  ) |>
  select(unique_id, entity)
ids_b <- df_b |>
  mutate(
    unique_id = row_number(),
    entity = str_extract(rec_id, '(?<=rec-)\\d+')
  ) |>
  select(unique_id, entity)

positives <- inner_join(ids_a, ids_b, by = 'entity', suffix = c('_l', '_r')) |>
  mutate(is_match = 1L) |>
  select(unique_id_l, unique_id_r, is_match)

candidate_pairs <- predict(model, threshold = 0) |>
  select(unique_id_l, unique_id_r) |>
  distinct()

set.seed(42)
candidate_negatives <- candidate_pairs |>
  left_join(ids_a, by = c('unique_id_l' = 'unique_id')) |>
  left_join(
    ids_b,
    by = c('unique_id_r' = 'unique_id'),
    suffix = c('_l', '_r')
  ) |>
  filter(entity_l != entity_r)

negatives <- candidate_negatives |>
  slice_sample(n = min(10000L, nrow(candidate_negatives))) |>
  transmute(unique_id_l, unique_id_r, is_match = 0L)

labels <- bind_rows(positives, negatives) |>
  distinct(unique_id_l, unique_id_r, .keep_all = TRUE)
labels
```

### Precision, recall, and F1

The accuracy chart shows how precision, recall, and F1 vary across all thresholds.
Not every true match is recoverable because some records are too corrupted, so the goal is balance rather than perfect recall.

```{r accuracy}
#| fig-width: 6
#| fig-height: 3.5
il_accuracy(model, labels = labels) |>
  autoplot()
```

### ROC curve

```{r roc}
#| fig-width: 5
#| fig-height: 4
il_roc(model, labels = labels) |>
  autoplot()
```

### Precision-recall curve

```{r pr}
#| fig-width: 5
#| fig-height: 4
il_precision_recall(model, labels = labels) |>
  autoplot()
```

### Errors at threshold 0.9

False negatives are typically heavily corrupted records where key fields are missing or garbled beyond what the comparison thresholds can accommodate.
False positives are pairs that coincidentally agree on enough fields to exceed the threshold despite referring to different people.

```{r errors}
errs <- il_errors(model, labels = labels, threshold = 0.9)
errs |>
  filter(error_type == 'false_negative')
errs |>
  filter(error_type == 'false_positive')
```

## Cleanup

```{r cleanup}
il_cleanup(model)
il_cleanup_all(con)
DBI::dbDisconnect(con, shutdown = TRUE)
```
