---
title: "Scaling and Accuracy with `pseudopeople`"
format:
  typst:
    syntax-highlighting: idiomatic
    keep-typ: true
---

This document benchmarks `irelink` on five `pseudopeople` scenarios using the parquet files in the sibling `pseudopeople-ri` repository.
Every scenario reads the parquet files directly through DuckDB, so the full source tables stay out of R memory.

The same comparison specification and the same three blocking rules are used across all five tests:

1. `first_name + last_name`
2. `last_name + date_of_birth`
3. `zipcode + date_of_birth`

```{r setup}
devtools::load_all()
library(dplyr)
library(ggplot2)
library(knitr)
library(tictoc)

pseudopeople_dir <- '../../../pseudopeople-ri'

zero_path <- paste0(pseudopeople_dir, '/psp_census_2020_zero_noise.parquet')
default_path <- paste0(
  pseudopeople_dir,
  '/psp_census_2020_default_noise.parquet'
)
high_path <- paste0(pseudopeople_dir, '/psp_census_2020_high_noise.parquet')

common_spec <- il_spec() |>
  il_compare(first_name, cl_name()) |>
  il_compare(last_name, cl_name()) |>
  il_compare(date_of_birth, cl_damerau_levenshtein(1, 2)) |>
  il_compare(street_number, cl_damerau_levenshtein(1)) |>
  il_compare(city, cl_exact()) |>
  il_compare(zipcode, cl_damerau_levenshtein(1)) |>
  il_block_on(first_name, last_name) |>
  il_block_on(last_name, date_of_birth) |>
  il_block_on(zipcode, date_of_birth)
```

## No noise to zero

```{r zero-zero-data}
zero_zero_con <- DBI::dbConnect(duckdb::duckdb())

DBI::dbExecute(
  zero_zero_con,
  paste0(
    'CREATE VIEW zero_zero_left AS ',
    "SELECT * FROM read_parquet('",
    zero_path,
    "')"
  )
)
DBI::dbExecute(
  zero_zero_con,
  paste0(
    'CREATE VIEW zero_zero_right AS ',
    "SELECT * FROM read_parquet('",
    zero_path,
    "')"
  )
)

zero_zero_left <- tbl(zero_zero_con, 'zero_zero_left')
zero_zero_right <- tbl(zero_zero_con, 'zero_zero_right')
```

### Blocking

```{r zero-zero-blocking-data}
tic.clearlog()
tic('zero_zero_blocking', quiet = TRUE)
zero_zero_blocking <- il_count_pairs(
  zero_zero_left,
  zero_zero_right,
  block_on(first_name, last_name),
  block_on(last_name, date_of_birth),
  block_on(zipcode, date_of_birth),
  con = zero_zero_con,
  link_type = 'link'
)
zero_zero_blocking_timing <- toc(log = TRUE, quiet = TRUE)
zero_zero_blocking_time <- as.numeric(
  zero_zero_blocking_timing$toc - zero_zero_blocking_timing$tic
)
zero_zero_candidate_pairs <- max(zero_zero_blocking$cumulative_pairs)
zero_zero_cartesian_share <- max(zero_zero_blocking$pct_of_cartesian)
```

This is the clean ceiling case: the same zero-noise census file appears on both sides, so the benchmark should be close to a keyed merge.
The blocking rules reduce the search to `r format(round(zero_zero_candidate_pairs), big.mark = ",", scientific = FALSE)` candidate pairs, which is `r round(100 * zero_zero_cartesian_share, 3)`% of the full cross-product.

```{r zero-zero-blocking-plot}
#| fig-width: 7
#| fig-height: 3.2
autoplot(zero_zero_blocking)
```

### Model fit

```{r zero-zero-model}
tic.clearlog()
tic('zero_zero_fit', quiet = TRUE)
zero_zero_model <- il_model(
  zero_zero_left,
  zero_zero_right,
  spec = common_spec,
  con = zero_zero_con,
  link_type = 'link'
) |>
  il_estimate_prior(
    block_on(first_name, last_name, date_of_birth),
    recall = 0.8
  ) |>
  il_estimate_u(max_pairs = 5e5) |>
  il_estimate_em(
    blocking = block_on(first_name, last_name),
    max_iterations = 8L
  ) |>
  il_estimate_em(
    blocking = block_on(zipcode, date_of_birth),
    max_iterations = 8L
  )
zero_zero_fit_timing <- toc(log = TRUE, quiet = TRUE)
zero_zero_fit_time <- as.numeric(
  zero_zero_fit_timing$toc - zero_zero_fit_timing$tic
)
```

The model fit includes prior estimation, u-estimation, and two EM sessions.
This step takes `r round(zero_zero_fit_time, 1)` seconds in the clean linkage ceiling case.

```{r zero-zero-model-plot}
#| fig-width: 7
#| fig-height: 4
autoplot(zero_zero_model, type = 'parameters')
```

```{r zero-zero-history}
#| fig-width: 7
#| fig-height: 4
il_training_history(zero_zero_model) |>
  autoplot()
```

### Accuracy

```{r zero-zero-accuracy-data}
tic.clearlog()
tic('zero_zero_accuracy', quiet = TRUE)
zero_zero_accuracy <- il_accuracy(
  zero_zero_model,
  labels_col = 'simulant_id'
)
zero_zero_accuracy_timing <- toc(log = TRUE, quiet = TRUE)
zero_zero_accuracy_time <- as.numeric(
  zero_zero_accuracy_timing$toc - zero_zero_accuracy_timing$tic
)
zero_zero_best <- zero_zero_accuracy |>
  slice_max(f1, n = 1, with_ties = FALSE)
```

At the best threshold (`r round(zero_zero_best$threshold, 3)`), precision is `r round(100 * zero_zero_best$precision, 2)`%, recall is `r round(100 * zero_zero_best$recall, 2)`%, and F1 is `r round(zero_zero_best$f1, 3)`.
Accuracy evaluation takes `r round(zero_zero_accuracy_time, 1)` seconds.

```{r zero-zero-accuracy-plot}
#| fig-width: 6.5
#| fig-height: 3.8
autoplot(zero_zero_accuracy)
```

### Prediction

```{r zero-zero-prediction-data}
tic.clearlog()
tic('zero_zero_prediction', quiet = TRUE)
zero_zero_pairs <- predict(
  zero_zero_model,
  threshold = zero_zero_best$threshold
)
zero_zero_prediction_timing <- toc(log = TRUE, quiet = TRUE)
zero_zero_prediction_time <- as.numeric(
  zero_zero_prediction_timing$toc - zero_zero_prediction_timing$tic
)
zero_zero_predicted_links <- nrow(zero_zero_pairs)

tic.clearlog()
tic('zero_zero_prediction_correctness', quiet = TRUE)
zero_zero_confusion <- il_confusion_matrix(
  zero_zero_model,
  threshold = zero_zero_best$threshold,
  labels_col = 'simulant_id'
)
zero_zero_prediction_correctness_timing <- toc(log = TRUE, quiet = TRUE)
zero_zero_prediction_correctness_time <- as.numeric(
  zero_zero_prediction_correctness_timing$toc -
    zero_zero_prediction_correctness_timing$tic
)
```

At the best F1 threshold, `predict()` returns `r format(zero_zero_predicted_links, big.mark = ",", scientific = FALSE)` links in `r round(zero_zero_prediction_time, 1)` seconds.
Thresholded prediction correctness is precision `r round(100 * zero_zero_confusion$precision, 2)`%, recall `r round(100 * zero_zero_confusion$recall, 2)`%, and F1 `r round(zero_zero_confusion$f1, 3)`.

### Timing

```{r zero-zero-timings}
zero_zero_timings <- tibble::tibble(
  stage = c(
    'Blocking analysis',
    'Model fitting',
    'Accuracy evaluation',
    'Prediction',
    'Prediction correctness',
    'Total'
  ),
  seconds = c(
    zero_zero_blocking_time,
    zero_zero_fit_time,
    zero_zero_accuracy_time,
    zero_zero_prediction_time,
    zero_zero_prediction_correctness_time,
    zero_zero_blocking_time +
      zero_zero_fit_time +
      zero_zero_accuracy_time +
      zero_zero_prediction_time +
      zero_zero_prediction_correctness_time
  )
)

zero_zero_timings
```

```{r zero-zero-cleanup}
#| include: false
il_cleanup(zero_zero_model)
il_cleanup_all(zero_zero_con)
DBI::dbDisconnect(zero_zero_con, shutdown = TRUE)
```

## No noise to default

```{r zero-default-data}
zero_default_con <- DBI::dbConnect(duckdb::duckdb())

DBI::dbExecute(
  zero_default_con,
  paste0(
    'CREATE VIEW zero_default_left AS ',
    "SELECT * FROM read_parquet('",
    zero_path,
    "')"
  )
)
DBI::dbExecute(
  zero_default_con,
  paste0(
    'CREATE VIEW zero_default_right AS ',
    "SELECT * FROM read_parquet('",
    default_path,
    "')"
  )
)

zero_default_left <- tbl(zero_default_con, 'zero_default_left')
zero_default_right <- tbl(zero_default_con, 'zero_default_right')
```

### Blocking

```{r zero-default-blocking-data}
tic.clearlog()
tic('zero_default_blocking', quiet = TRUE)
zero_default_blocking <- il_count_pairs(
  zero_default_left,
  zero_default_right,
  block_on(first_name, last_name),
  block_on(last_name, date_of_birth),
  block_on(zipcode, date_of_birth),
  con = zero_default_con,
  link_type = 'link'
)
zero_default_blocking_timing <- toc(log = TRUE, quiet = TRUE)
zero_default_blocking_time <- as.numeric(
  zero_default_blocking_timing$toc - zero_default_blocking_timing$tic
)
zero_default_candidate_pairs <- max(zero_default_blocking$cumulative_pairs)
zero_default_cartesian_share <- max(zero_default_blocking$pct_of_cartesian)
```

This links the clean census extract to the default-noise administrative extract.
The blocking rules keep the candidate space to `r format(round(zero_default_candidate_pairs), big.mark = ",", scientific = FALSE)` pairs, or `r round(100 * zero_default_cartesian_share, 3)`% of the full cross-product.

```{r zero-default-blocking-plot}
#| fig-width: 7
#| fig-height: 3.2
autoplot(zero_default_blocking)
```

### Model fit

```{r zero-default-model}
tic.clearlog()
tic('zero_default_fit', quiet = TRUE)
zero_default_model <- il_model(
  zero_default_left,
  zero_default_right,
  spec = common_spec,
  con = zero_default_con,
  link_type = 'link'
) |>
  il_estimate_prior(
    block_on(first_name, last_name, date_of_birth),
    recall = 0.8
  ) |>
  il_estimate_u(max_pairs = 5e5) |>
  il_estimate_em(
    blocking = block_on(first_name, last_name),
    max_iterations = 8L
  ) |>
  il_estimate_em(
    blocking = block_on(zipcode, date_of_birth),
    max_iterations = 8L
  )
zero_default_fit_timing <- toc(log = TRUE, quiet = TRUE)
zero_default_fit_time <- as.numeric(
  zero_default_fit_timing$toc - zero_default_fit_timing$tic
)
```

This fit step takes `r round(zero_default_fit_time, 1)` seconds.
The same specification is used here, but the default-noise file makes the EM updates materially less clean than in the no-noise-to-zero case.

```{r zero-default-model-plot}
#| fig-width: 7
#| fig-height: 4
autoplot(zero_default_model, type = 'parameters')
```

```{r zero-default-history}
#| fig-width: 7
#| fig-height: 4
il_training_history(zero_default_model) |>
  autoplot()
```

### Accuracy

```{r zero-default-accuracy-data}
tic.clearlog()
tic('zero_default_accuracy', quiet = TRUE)
zero_default_accuracy <- il_accuracy(
  zero_default_model,
  labels_col = 'simulant_id'
)
zero_default_accuracy_timing <- toc(log = TRUE, quiet = TRUE)
zero_default_accuracy_time <- as.numeric(
  zero_default_accuracy_timing$toc - zero_default_accuracy_timing$tic
)
zero_default_best <- zero_default_accuracy |>
  slice_max(f1, n = 1, with_ties = FALSE)
```

At the best threshold (`r round(zero_default_best$threshold, 3)`), precision is `r round(100 * zero_default_best$precision, 2)`%, recall is `r round(100 * zero_default_best$recall, 2)`%, and F1 is `r round(zero_default_best$f1, 3)`.
Accuracy evaluation takes `r round(zero_default_accuracy_time, 1)` seconds.

```{r zero-default-accuracy-plot}
#| fig-width: 6.5
#| fig-height: 3.8
autoplot(zero_default_accuracy)
```

### Prediction

```{r zero-default-prediction-data}
tic.clearlog()
tic('zero_default_prediction', quiet = TRUE)
zero_default_pairs <- predict(
  zero_default_model,
  threshold = zero_default_best$threshold
)
zero_default_prediction_timing <- toc(log = TRUE, quiet = TRUE)
zero_default_prediction_time <- as.numeric(
  zero_default_prediction_timing$toc - zero_default_prediction_timing$tic
)
zero_default_predicted_links <- nrow(zero_default_pairs)

tic.clearlog()
tic('zero_default_prediction_correctness', quiet = TRUE)
zero_default_confusion <- il_confusion_matrix(
  zero_default_model,
  threshold = zero_default_best$threshold,
  labels_col = 'simulant_id'
)
zero_default_prediction_correctness_timing <- toc(log = TRUE, quiet = TRUE)
zero_default_prediction_correctness_time <- as.numeric(
  zero_default_prediction_correctness_timing$toc -
    zero_default_prediction_correctness_timing$tic
)
```

At the best F1 threshold, `predict()` returns `r format(zero_default_predicted_links, big.mark = ",", scientific = FALSE)` links in `r round(zero_default_prediction_time, 1)` seconds.
Thresholded prediction correctness is precision `r round(100 * zero_default_confusion$precision, 2)`%, recall `r round(100 * zero_default_confusion$recall, 2)`%, and F1 `r round(zero_default_confusion$f1, 3)`.

### Timing

```{r zero-default-timings}
zero_default_timings <- tibble::tibble(
  stage = c(
    'Blocking analysis',
    'Model fitting',
    'Accuracy evaluation',
    'Prediction',
    'Prediction correctness',
    'Total'
  ),
  seconds = c(
    zero_default_blocking_time,
    zero_default_fit_time,
    zero_default_accuracy_time,
    zero_default_prediction_time,
    zero_default_prediction_correctness_time,
    zero_default_blocking_time +
      zero_default_fit_time +
      zero_default_accuracy_time +
      zero_default_prediction_time +
      zero_default_prediction_correctness_time
  )
)

zero_default_timings
```

```{r zero-default-cleanup}
#| include: false
il_cleanup(zero_default_model)
il_cleanup_all(zero_default_con)
DBI::dbDisconnect(zero_default_con, shutdown = TRUE)
```

## No noise to high

```{r zero-high-data}
zero_high_con <- DBI::dbConnect(duckdb::duckdb())

DBI::dbExecute(
  zero_high_con,
  paste0(
    'CREATE VIEW zero_high_left AS ',
    "SELECT * FROM read_parquet('",
    zero_path,
    "')"
  )
)
DBI::dbExecute(
  zero_high_con,
  paste0(
    'CREATE VIEW zero_high_right AS ',
    "SELECT * FROM read_parquet('",
    high_path,
    "')"
  )
)

zero_high_left <- tbl(zero_high_con, 'zero_high_left')
zero_high_right <- tbl(zero_high_con, 'zero_high_right')
```

### Blocking

```{r zero-high-blocking-data}
tic.clearlog()
tic('zero_high_blocking', quiet = TRUE)
zero_high_blocking <- il_count_pairs(
  zero_high_left,
  zero_high_right,
  block_on(first_name, last_name),
  block_on(last_name, date_of_birth),
  block_on(zipcode, date_of_birth),
  con = zero_high_con,
  link_type = 'link'
)
zero_high_blocking_timing <- toc(log = TRUE, quiet = TRUE)
zero_high_blocking_time <- as.numeric(
  zero_high_blocking_timing$toc - zero_high_blocking_timing$tic
)
zero_high_candidate_pairs <- max(zero_high_blocking$cumulative_pairs)
zero_high_cartesian_share <- max(zero_high_blocking$pct_of_cartesian)
```

This links the clean census extract to the high-noise administrative extract.
The candidate universe is `r format(round(zero_high_candidate_pairs), big.mark = ",", scientific = FALSE)` pairs, or `r round(100 * zero_high_cartesian_share, 3)`% of the full cross-product.

```{r zero-high-blocking-plot}
#| fig-width: 7
#| fig-height: 3.2
autoplot(zero_high_blocking)
```

### Model fit

```{r zero-high-model}
tic.clearlog()
tic('zero_high_fit', quiet = TRUE)
zero_high_model <- il_model(
  zero_high_left,
  zero_high_right,
  spec = common_spec,
  con = zero_high_con,
  link_type = 'link'
) |>
  il_estimate_prior(
    block_on(first_name, last_name, date_of_birth),
    recall = 0.8
  ) |>
  il_estimate_u(max_pairs = 5e5) |>
  il_estimate_em(
    blocking = block_on(first_name, last_name),
    max_iterations = 8L
  ) |>
  il_estimate_em(
    blocking = block_on(zipcode, date_of_birth),
    max_iterations = 8L
  )
zero_high_fit_timing <- toc(log = TRUE, quiet = TRUE)
zero_high_fit_time <- as.numeric(
  zero_high_fit_timing$toc - zero_high_fit_timing$tic
)
```

The fit step takes `r round(zero_high_fit_time, 1)` seconds.
This is the hardest one-sided linkage task in the document, so the fitted parameter separation is weaker than in the no-noise-to-default case.

```{r zero-high-model-plot}
#| fig-width: 7
#| fig-height: 4
autoplot(zero_high_model, type = 'parameters')
```

```{r zero-high-history}
#| fig-width: 7
#| fig-height: 4
il_training_history(zero_high_model) |>
  autoplot()
```

### Accuracy

```{r zero-high-accuracy-data}
tic.clearlog()
tic('zero_high_accuracy', quiet = TRUE)
zero_high_accuracy <- il_accuracy(
  zero_high_model,
  labels_col = 'simulant_id'
)
zero_high_accuracy_timing <- toc(log = TRUE, quiet = TRUE)
zero_high_accuracy_time <- as.numeric(
  zero_high_accuracy_timing$toc - zero_high_accuracy_timing$tic
)
zero_high_best <- zero_high_accuracy |>
  slice_max(f1, n = 1, with_ties = FALSE)
```

At the best threshold (`r round(zero_high_best$threshold, 3)`), precision is `r round(100 * zero_high_best$precision, 2)`%, recall is `r round(100 * zero_high_best$recall, 2)`%, and F1 is `r round(zero_high_best$f1, 3)`.
Accuracy evaluation takes `r round(zero_high_accuracy_time, 1)` seconds.

```{r zero-high-accuracy-plot}
#| fig-width: 6.5
#| fig-height: 3.8
autoplot(zero_high_accuracy)
```

### Prediction

```{r zero-high-prediction-data}
tic.clearlog()
tic('zero_high_prediction', quiet = TRUE)
zero_high_pairs <- predict(
  zero_high_model,
  threshold = zero_high_best$threshold
)
zero_high_prediction_timing <- toc(log = TRUE, quiet = TRUE)
zero_high_prediction_time <- as.numeric(
  zero_high_prediction_timing$toc - zero_high_prediction_timing$tic
)
zero_high_predicted_links <- nrow(zero_high_pairs)

tic.clearlog()
tic('zero_high_prediction_correctness', quiet = TRUE)
zero_high_confusion <- il_confusion_matrix(
  zero_high_model,
  threshold = zero_high_best$threshold,
  labels_col = 'simulant_id'
)
zero_high_prediction_correctness_timing <- toc(log = TRUE, quiet = TRUE)
zero_high_prediction_correctness_time <- as.numeric(
  zero_high_prediction_correctness_timing$toc -
    zero_high_prediction_correctness_timing$tic
)
```

At the best F1 threshold, `predict()` returns `r format(zero_high_predicted_links, big.mark = ",", scientific = FALSE)` links in `r round(zero_high_prediction_time, 1)` seconds.
Thresholded prediction correctness is precision `r round(100 * zero_high_confusion$precision, 2)`%, recall `r round(100 * zero_high_confusion$recall, 2)`%, and F1 `r round(zero_high_confusion$f1, 3)`.

### Timing

```{r zero-high-timings}
zero_high_timings <- tibble::tibble(
  stage = c(
    'Blocking analysis',
    'Model fitting',
    'Accuracy evaluation',
    'Prediction',
    'Prediction correctness',
    'Total'
  ),
  seconds = c(
    zero_high_blocking_time,
    zero_high_fit_time,
    zero_high_accuracy_time,
    zero_high_prediction_time,
    zero_high_prediction_correctness_time,
    zero_high_blocking_time +
      zero_high_fit_time +
      zero_high_accuracy_time +
      zero_high_prediction_time +
      zero_high_prediction_correctness_time
  )
)

zero_high_timings
```

```{r zero-high-cleanup}
#| include: false
il_cleanup(zero_high_model)
il_cleanup_all(zero_high_con)
DBI::dbDisconnect(zero_high_con, shutdown = TRUE)
```

## Default noise to high

```{r default-high-data}
default_high_con <- DBI::dbConnect(duckdb::duckdb())

DBI::dbExecute(
  default_high_con,
  paste0(
    'CREATE VIEW default_high_left AS ',
    "SELECT * FROM read_parquet('",
    default_path,
    "')"
  )
)
DBI::dbExecute(
  default_high_con,
  paste0(
    'CREATE VIEW default_high_right AS ',
    "SELECT * FROM read_parquet('",
    high_path,
    "')"
  )
)

default_high_left <- tbl(default_high_con, 'default_high_left')
default_high_right <- tbl(default_high_con, 'default_high_right')
```

### Blocking

```{r default-high-blocking-data}
tic.clearlog()
tic('default_high_blocking', quiet = TRUE)
default_high_blocking <- il_count_pairs(
  default_high_left,
  default_high_right,
  block_on(first_name, last_name),
  block_on(last_name, date_of_birth),
  block_on(zipcode, date_of_birth),
  con = default_high_con,
  link_type = 'link'
)
default_high_blocking_timing <- toc(log = TRUE, quiet = TRUE)
default_high_blocking_time <- as.numeric(
  default_high_blocking_timing$toc - default_high_blocking_timing$tic
)
default_high_candidate_pairs <- max(default_high_blocking$cumulative_pairs)
default_high_cartesian_share <- max(default_high_blocking$pct_of_cartesian)
```

This is the realistic administrative linkage case, with noise on both sides.
The candidate universe is `r format(round(default_high_candidate_pairs), big.mark = ",", scientific = FALSE)` pairs, or `r round(100 * default_high_cartesian_share, 3)`% of the full cross-product.

```{r default-high-blocking-plot}
#| fig-width: 7
#| fig-height: 3.2
autoplot(default_high_blocking)
```

### Model fit

```{r default-high-model}
tic.clearlog()
tic('default_high_fit', quiet = TRUE)
default_high_model <- il_model(
  default_high_left,
  default_high_right,
  spec = common_spec,
  con = default_high_con,
  link_type = 'link'
) |>
  il_estimate_prior(
    block_on(first_name, last_name, date_of_birth),
    recall = 0.8
  ) |>
  il_estimate_u(max_pairs = 5e5) |>
  il_estimate_em(
    blocking = block_on(first_name, last_name),
    max_iterations = 8L
  ) |>
  il_estimate_em(
    blocking = block_on(zipcode, date_of_birth),
    max_iterations = 8L
  )
default_high_fit_timing <- toc(log = TRUE, quiet = TRUE)
default_high_fit_time <- as.numeric(
  default_high_fit_timing$toc - default_high_fit_timing$tic
)
```

The fit step takes `r round(default_high_fit_time, 1)` seconds.
Because both sides are noisy, this is the most realistic two-file linkage benchmark in the set.

```{r default-high-model-plot}
#| fig-width: 7
#| fig-height: 4
autoplot(default_high_model, type = 'parameters')
```

```{r default-high-history}
#| fig-width: 7
#| fig-height: 4
il_training_history(default_high_model) |>
  autoplot()
```

### Accuracy

```{r default-high-accuracy-data}
tic.clearlog()
tic('default_high_accuracy', quiet = TRUE)
default_high_accuracy <- il_accuracy(
  default_high_model,
  labels_col = 'simulant_id'
)
default_high_accuracy_timing <- toc(log = TRUE, quiet = TRUE)
default_high_accuracy_time <- as.numeric(
  default_high_accuracy_timing$toc - default_high_accuracy_timing$tic
)
default_high_best <- default_high_accuracy |>
  slice_max(f1, n = 1, with_ties = FALSE)
```

At the best threshold (`r round(default_high_best$threshold, 3)`), precision is `r round(100 * default_high_best$precision, 2)`%, recall is `r round(100 * default_high_best$recall, 2)`%, and F1 is `r round(default_high_best$f1, 3)`.
Accuracy evaluation takes `r round(default_high_accuracy_time, 1)` seconds.

```{r default-high-accuracy-plot}
#| fig-width: 6.5
#| fig-height: 3.8
autoplot(default_high_accuracy)
```

### Prediction

```{r default-high-prediction-data}
tic.clearlog()
tic('default_high_prediction', quiet = TRUE)
default_high_pairs <- predict(
  default_high_model,
  threshold = default_high_best$threshold
)
default_high_prediction_timing <- toc(log = TRUE, quiet = TRUE)
default_high_prediction_time <- as.numeric(
  default_high_prediction_timing$toc - default_high_prediction_timing$tic
)
default_high_predicted_links <- nrow(default_high_pairs)

tic.clearlog()
tic('default_high_prediction_correctness', quiet = TRUE)
default_high_confusion <- il_confusion_matrix(
  default_high_model,
  threshold = default_high_best$threshold,
  labels_col = 'simulant_id'
)
default_high_prediction_correctness_timing <- toc(log = TRUE, quiet = TRUE)
default_high_prediction_correctness_time <- as.numeric(
  default_high_prediction_correctness_timing$toc -
    default_high_prediction_correctness_timing$tic
)
```

At the best F1 threshold, `predict()` returns `r format(default_high_predicted_links, big.mark = ",", scientific = FALSE)` links in `r round(default_high_prediction_time, 1)` seconds.
Thresholded prediction correctness is precision `r round(100 * default_high_confusion$precision, 2)`%, recall `r round(100 * default_high_confusion$recall, 2)`%, and F1 `r round(default_high_confusion$f1, 3)`.

### Timing

```{r default-high-timings}
default_high_timings <- tibble::tibble(
  stage = c(
    'Blocking analysis',
    'Model fitting',
    'Accuracy evaluation',
    'Prediction',
    'Prediction correctness',
    'Total'
  ),
  seconds = c(
    default_high_blocking_time,
    default_high_fit_time,
    default_high_accuracy_time,
    default_high_prediction_time,
    default_high_prediction_correctness_time,
    default_high_blocking_time +
      default_high_fit_time +
      default_high_accuracy_time +
      default_high_prediction_time +
      default_high_prediction_correctness_time
  )
)

default_high_timings
```

```{r default-high-cleanup}
#| include: false
il_cleanup(default_high_model)
il_cleanup_all(default_high_con)
DBI::dbDisconnect(default_high_con, shutdown = TRUE)
```

## Stacked zero, default, and high noise

```{r stacked-dedupe-data}
stacked_dedupe_con <- DBI::dbConnect(duckdb::duckdb())

DBI::dbExecute(
  stacked_dedupe_con,
  paste0(
    'CREATE VIEW stacked_dedupe_input AS ',
    "SELECT 'zero' AS source_file, * FROM read_parquet('",
    zero_path,
    "') ",
    'UNION ALL ',
    "SELECT 'default' AS source_file, * FROM read_parquet('",
    default_path,
    "') ",
    'UNION ALL ',
    "SELECT 'high' AS source_file, * FROM read_parquet('",
    high_path,
    "')"
  )
)

stacked_dedupe_input <- tbl(stacked_dedupe_con, 'stacked_dedupe_input')
```

### Blocking

```{r stacked-dedupe-blocking-data}
tic.clearlog()
tic('stacked_dedupe_blocking', quiet = TRUE)
stacked_dedupe_blocking <- il_count_pairs(
  stacked_dedupe_input,
  block_on(first_name, last_name),
  block_on(last_name, date_of_birth),
  block_on(zipcode, date_of_birth),
  con = stacked_dedupe_con,
  link_type = 'dedupe'
)
stacked_dedupe_blocking_timing <- toc(log = TRUE, quiet = TRUE)
stacked_dedupe_blocking_time <- as.numeric(
  stacked_dedupe_blocking_timing$toc - stacked_dedupe_blocking_timing$tic
)
stacked_dedupe_candidate_pairs <- max(stacked_dedupe_blocking$cumulative_pairs)
stacked_dedupe_cartesian_share <- max(stacked_dedupe_blocking$pct_of_cartesian)
```

This stacks the clean, default-noise, and high-noise census extracts and deduplicates the combined table.
The blocking rules cut the search to `r format(round(stacked_dedupe_candidate_pairs), big.mark = ",", scientific = FALSE)` candidate pairs, or `r round(100 * stacked_dedupe_cartesian_share, 3)`% of all within-table pairs.

```{r stacked-dedupe-blocking-plot}
#| fig-width: 7
#| fig-height: 3.2
autoplot(stacked_dedupe_blocking)
```

### Model fit

```{r stacked-dedupe-model}
tic.clearlog()
tic('stacked_dedupe_fit', quiet = TRUE)
stacked_dedupe_model <- il_model(
  stacked_dedupe_input,
  spec = common_spec,
  con = stacked_dedupe_con,
  link_type = 'dedupe'
) |>
  il_estimate_prior(
    block_on(first_name, last_name, date_of_birth),
    recall = 0.8
  ) |>
  il_estimate_u(max_pairs = 5e5) |>
  il_estimate_em(
    blocking = block_on(first_name, last_name),
    max_iterations = 8L
  ) |>
  il_estimate_em(
    blocking = block_on(zipcode, date_of_birth),
    max_iterations = 8L
  )
stacked_dedupe_fit_timing <- toc(log = TRUE, quiet = TRUE)
stacked_dedupe_fit_time <- as.numeric(
  stacked_dedupe_fit_timing$toc - stacked_dedupe_fit_timing$tic
)
```

This fit step takes `r round(stacked_dedupe_fit_time, 1)` seconds.
The stacked deduplication case is the heaviest benchmark because it includes cross-source duplicates and the small number of within-source duplicates present in the noisy files.

```{r stacked-dedupe-model-plot}
#| fig-width: 7
#| fig-height: 4
autoplot(stacked_dedupe_model, type = 'parameters')
```

```{r stacked-dedupe-history}
#| fig-width: 7
#| fig-height: 4
il_training_history(stacked_dedupe_model) |>
  autoplot()
```

### Accuracy

```{r stacked-dedupe-accuracy-data}
tic.clearlog()
tic('stacked_dedupe_accuracy', quiet = TRUE)
stacked_dedupe_accuracy <- il_accuracy(
  stacked_dedupe_model,
  labels_col = 'simulant_id'
)
stacked_dedupe_accuracy_timing <- toc(log = TRUE, quiet = TRUE)
stacked_dedupe_accuracy_time <- as.numeric(
  stacked_dedupe_accuracy_timing$toc - stacked_dedupe_accuracy_timing$tic
)
stacked_dedupe_best <- stacked_dedupe_accuracy |>
  slice_max(f1, n = 1, with_ties = FALSE)
```

At the best threshold (`r round(stacked_dedupe_best$threshold, 3)`), precision is `r round(100 * stacked_dedupe_best$precision, 2)`%, recall is `r round(100 * stacked_dedupe_best$recall, 2)`%, and F1 is `r round(stacked_dedupe_best$f1, 3)`.
Accuracy evaluation takes `r round(stacked_dedupe_accuracy_time, 1)` seconds.

```{r stacked-dedupe-accuracy-plot}
#| fig-width: 6.5
#| fig-height: 3.8
autoplot(stacked_dedupe_accuracy)
```

### Prediction

```{r stacked-dedupe-prediction-data}
tic.clearlog()
tic('stacked_dedupe_prediction', quiet = TRUE)
stacked_dedupe_pairs <- predict(
  stacked_dedupe_model,
  threshold = stacked_dedupe_best$threshold
)
stacked_dedupe_prediction_timing <- toc(log = TRUE, quiet = TRUE)
stacked_dedupe_prediction_time <- as.numeric(
  stacked_dedupe_prediction_timing$toc - stacked_dedupe_prediction_timing$tic
)
stacked_dedupe_predicted_pairs <- nrow(stacked_dedupe_pairs)

tic.clearlog()
tic('stacked_dedupe_prediction_correctness', quiet = TRUE)
stacked_dedupe_confusion <- il_confusion_matrix(
  stacked_dedupe_model,
  threshold = stacked_dedupe_best$threshold,
  labels_col = 'simulant_id'
)
stacked_dedupe_prediction_correctness_timing <- toc(log = TRUE, quiet = TRUE)
stacked_dedupe_prediction_correctness_time <- as.numeric(
  stacked_dedupe_prediction_correctness_timing$toc -
    stacked_dedupe_prediction_correctness_timing$tic
)
```

At the best F1 threshold, `predict()` returns `r format(stacked_dedupe_predicted_pairs, big.mark = ",", scientific = FALSE)` scored duplicate pairs in `r round(stacked_dedupe_prediction_time, 1)` seconds.
Pairwise correctness at that threshold is precision `r round(100 * stacked_dedupe_confusion$precision, 2)`%, recall `r round(100 * stacked_dedupe_confusion$recall, 2)`%, and F1 `r round(stacked_dedupe_confusion$f1, 3)`.

### Clustering

```{r stacked-dedupe-clustering-data}
tic.clearlog()
tic('stacked_dedupe_clustering', quiet = TRUE)
stacked_dedupe_clusters <- il_cluster(stacked_dedupe_pairs)
stacked_dedupe_clustering_timing <- toc(log = TRUE, quiet = TRUE)
stacked_dedupe_clustering_time <- as.numeric(
  stacked_dedupe_clustering_timing$toc - stacked_dedupe_clustering_timing$tic
)
stacked_dedupe_cluster_sizes <- stacked_dedupe_clusters |>
  count(cluster_id, name = 'cluster_size', sort = TRUE)
stacked_dedupe_n_clusters <- nrow(stacked_dedupe_cluster_sizes)
stacked_dedupe_largest_cluster <- if (nrow(stacked_dedupe_cluster_sizes) > 0) {
  max(stacked_dedupe_cluster_sizes$cluster_size)
} else {
  0L
}

tic.clearlog()
tic('stacked_dedupe_cluster_correctness', quiet = TRUE)
stacked_dedupe_cluster_confusion <- il_cluster_confusion_matrix(
  stacked_dedupe_model,
  labels_col = 'simulant_id',
  threshold = stacked_dedupe_best$threshold
)
stacked_dedupe_cluster_correctness_timing <- toc(log = TRUE, quiet = TRUE)
stacked_dedupe_cluster_correctness_time <- as.numeric(
  stacked_dedupe_cluster_correctness_timing$toc -
    stacked_dedupe_cluster_correctness_timing$tic
)
```

Connected-components clustering yields `r format(stacked_dedupe_n_clusters, big.mark = ",", scientific = FALSE)` clusters, with largest cluster size `r format(stacked_dedupe_largest_cluster, big.mark = ",", scientific = FALSE)`, in `r round(stacked_dedupe_clustering_time, 1)` seconds.
Cluster-level duplicate detection correctness is precision `r round(100 * stacked_dedupe_cluster_confusion$precision, 2)`%, recall `r round(100 * stacked_dedupe_cluster_confusion$recall, 2)`%, and F1 `r round(stacked_dedupe_cluster_confusion$f1, 3)`.

### Timing

```{r stacked-dedupe-timings}
stacked_dedupe_timings <- tibble::tibble(
  stage = c(
    'Blocking analysis',
    'Model fitting',
    'Accuracy evaluation',
    'Prediction',
    'Prediction correctness',
    'Clustering',
    'Cluster correctness',
    'Total'
  ),
  seconds = c(
    stacked_dedupe_blocking_time,
    stacked_dedupe_fit_time,
    stacked_dedupe_accuracy_time,
    stacked_dedupe_prediction_time,
    stacked_dedupe_prediction_correctness_time,
    stacked_dedupe_clustering_time,
    stacked_dedupe_cluster_correctness_time,
    stacked_dedupe_blocking_time +
      stacked_dedupe_fit_time +
      stacked_dedupe_accuracy_time +
      stacked_dedupe_prediction_time +
      stacked_dedupe_prediction_correctness_time +
      stacked_dedupe_clustering_time +
      stacked_dedupe_cluster_correctness_time
  )
)

stacked_dedupe_timings
```

```{r stacked-dedupe-cleanup}
#| include: false
il_cleanup(stacked_dedupe_model)
il_cleanup_all(stacked_dedupe_con)
DBI::dbDisconnect(stacked_dedupe_con, shutdown = TRUE)
```
